Appendix: A fragile guardrail---The minimal effect of election fraud on protest intensity in non-democracies

### 2025-11-16 Output/authoritarian_only_pdf_pop_rev.Rmd
Final edits to discussion

The duration analysis consistently shows that more election manipulation is associated with quicker incumbent victories over protest movements, both independently and in conjunction with low incumbent vote-share. In real contests in which opposition and incumbent strengths are tested against each other, better manipulators reveal themselves to be stronger rather than weaker. This holds true even when incumbent vote-share is low, when the regime-weakness model would make the opposite prediction---in that model, incumbents who are unpopular and manipulate heavily ought to be the weakest and struggle hardest to quell protest. It is therefore inconsistent with Hypothesis 1, casting doubt on the 'regime-weakness' model of manipulation. Further, the relationship holds despite controlling for indicators of other possible explanations of the relationship between low fraud and long-lived protest, including civil society strength, media openness, regional electoral protest waves, number of protests and median protest size. This is consistent with Hypothesis 2, and supportive of what I have called the 'regime-strength' model of electoral manipulation.

This finding supports and extends the line of research initiated by Simpser [@simpser2013]. Understanding election manipulation as indicative of effective sources of regime strength comports with a growing literature on the principal-agent dynamics of manipulation [@Harvey2019;@rundlett2016deliver;@Harvey2022] and the importance of patronage resources and electoral processes in sustaining authoritarian regimes [@hale_patronal_2014]. This study offers empirical evidence that authoritarian governments who manipulate more are stronger than those who manipulate less (all else equal). In addition to offering a counter-argument to regime-weakness models, it also illustrates how the mechanisms which underpin Simpser's model of manipulation-as-strength---coordination effects among pro-regime agents and costly signaling of incumbent resources---can function even when such signals of strength have failed to 'deter' protest.

Specifically, the results highlight patterns that might be considered puzzling, whichever model of electoral manipulation we prefer. For the regime-weakness model, it is potentially puzzling that most manipulated elections go un-protested. Regime-weakness models may explain this pattern by arguing that, most of the time, incumbents produce the 'right' amount of manipulation to avoid protest in equilibrium. Such careful calibration is difficult to ensure, however, when incumbents must rely on thousands of low-level agents to produce the 'right' amount of manipulation; these agents may have collective incentives to shirk and underproduce, or to demonstrate their loyalty by over-producing. [ADD CITES] The results of this study add an empirical challenge, since it demonstrates that when the deterrence equilibrium fails (i.e. when the incumbent rigs elections and the opposition protests), higher levels of manipulation are associated with greater regime strength, not less.

In rejoinder, it might reasonably be wondered why protests occur after highly manipulated elections, if manipulation is a signal of strength? First, in my view, this interpretation unnecessarily exports the instrumental logic of protest from the regime-weakness model to the regime-strength model. As discussed earlier in the paper, instrumental calculations about overthrowing incumbents are but one component of a long and varied literature on protest, which has identified multiple other expressive and strategic motives for protest. These other motives may foster protest movements in the aftermath of rigged elections, even when severe electoral manipulation indicates considerable regime strength. Second, if manipulation does indeed indicate strength, incumbents have every incentive to build and deploy those strengths in reaction to stronger opposition movements (who will themselves be more capable of initiating protest). Evidence that incumbents do, indeed, appear to increase manipulation in the face of increased protest threat is presented in the supplementary appendix.

It also helps us make sense of secular patterns that are puzzling if election fraud indicates regime weakness, as others have held. The pattern depicted in Figure \@ref(fig:elex-hist) for example, which shows high levels of manipulation and low levels of protest is sensible in a world where the more powerful and durable incumbents produce greater levels of manipulation, and where incumbents who struggle to produce high levels of manipulation are ones who face resource scarcity, institutional barriers to impunity for their agents, and defections among erstwhile allies. It also helps explain why some incumbents manufacture their way to highly implausible margins of victory, reasoning that a demonstration of their extra-electoral power may suggest to allies and bystanders that the incumbent may survive protest despite a low natural popularity. Scenarios where incumbents use massive fraud to post margins of victory in the eighty and ninety percentage point range are harder to explain under a worldview where more manipulation indicates deeper regime weakness.

@@

What does all this tell us? First, these results may inform how researchers think about the causes of election manipulation. They suggest that the degree to which the opposition's capacity to organize street protest (and related factors) can limit election fraud have been overestimated. Consequently, theories of electoral manipulation that emphasize alternatives---such as resource availability [@greene_why_2007], principal-agent problems [@rundlett2016deliver], and legal risks to manipulators [@Harvey2022]---should be further developed, and may yield new policy interventions. In particular, it may change some current understandings of the role of election monitors, courts, and other third parties in upholding election integrity. As noted earlier, prior studies argue that these actors restrain manipulation by revealing information about fraud and intensifying protest risk. However, if more extensive manipulation is associated with more quickly subduing protest movements, it may be that other mechanisms are at work. For instance, it may be that election observation makes it harder for front-line agents to avoid personal consequences for engaging in manipulation [@Harvey2022].

By showing that the infrastructure of fraud can actually reduce the cost of post-election protest to the regime, this result poses a challenge to theory of democratization by election [@Schedler2013;@Morgenbesser2019;@edgellWhenWhereElections2018]. If the capacity to engage in widespread manipulation makes it easier to both tolerate the existence of an opposition and to suppress it during periods of contestation [@dahlPolyarchyParticipationOpposition1971], electoral authoritarianism is something of an institutional cul-de-sac: the capacity to generate severe election fraud makes it appealing for closed authoritarian regimes to introduce elections, while the manageable risk of protest makes it unappealing to further democratize.

@@

```
-->
Finally, this paper does not examine the role of protest risk in insulating established democracies from democratic backsliding [@fearon_self-enforcing_2011]. This question deserves further investigation, but it is possible to draw a few implications from this study. First, even if protest risk fails to serve as a deterrent in non-democracies, it may still do so in more democratic settings. In non-democracies, protest initiation is most likely when fraud is very high and incumbent vote-shares very low. The likelihood of protest initiation in a democracy may well be higher, triggering at lower levels of fraud and higher incumbent vote-shares, since opposition actors are likely to be better resourced in a liberal democracy (and the government more constrained). At the same time, if the theory described here is accurate, liberal democracies would be constrained against using large-scale election fraud even in the absence of protest risk: they lack the patronage resources and arbitrary enforcement of laws needed for such a scenario.

@@

In sum, the results indicate that we should be cautious when building and interpreting theories that emphasize the risk of mass protest as a trip-wire to be avoided in non-democracies, and add empirical evidence to Przeworski's [-@przeworskiFormalModelsAuthoritarian2022] critique of such models. The rarity of post-election protest, especially successful protest, suggests that such events are highly contingent; they hinge on a multitude of factors that go beyond the regime's efforts to bias election outcomes. These findings indicate that, while incumbents do seek to contain protest, more effective manipulators are also more capable of defeating protest movements.

These results are consistent with different specifications presented in an online appendix. In particular, the main results are supported when alternative measures of election integrity are used in place of the measure of election fraud. They are also supported when electoral democracies are included in the dataset. The relationship between fraud levels and incumbent defeat is also tested (along with incumbent ouster by protest). Additional robustness checks include models with data from election observer reports, with country fixed-effects, alternative specifications for selected control variables.

>>>>

The duration analysis consistently shows that more election manipulation is associated with quicker incumbent victories over protest movements, both independently and in conjunction with low incumbent vote-share. In real contests in which opposition and incumbent strengths are tested against each other, better manipulators reveal themselves to be stronger rather than weaker. This holds true even when incumbent vote-share is low, when the regime-weakness model would make the opposite prediction---in that model, incumbents who are unpopular and manipulate heavily ought to be the weakest and struggle hardest to quell protest. It is therefore inconsistent with Hypothesis 1, casting doubt on the 'regime-weakness' model of manipulation. Further, the relationship holds despite controlling for indicators of other possible explanations of the relationship between low fraud and long-lived protest, including civil society strength, media openness, regional electoral protest waves, number of protests and median protest size. While this study does not purport to show perfect causal identification, these robust correlations are presented as incompatible with the regime-weakness framework and supportive of the regime-strength model of manipulation.

Specifically, the results highlight patterns that might be considered puzzling, whichever model of electoral manipulation we prefer. For the regime-weakness model, it is potentially puzzling that most manipulated elections go un-protested. Regime-weakness models may explain this pattern by arguing that, most of the time, incumbents produce the 'right' amount of manipulation to avoid protest in equilibrium [@Little2012, e.g.]. Such careful calibration is difficult to ensure, however, when incumbents must rely on thousands of low-level agents to produce the 'right' amount of manipulation; these agents may have collective incentives to shirk and underproduce [@rundlett2016deliver], or to demonstrate their loyalty by over-producing [@choWhenAgentsDeliver2025]. The results of this study add an empirical challenge, since it demonstrates that when the deterrence equilibrium fails (i.e. when the incumbent rigs elections and the opposition protests), higher levels of manipulation are associated with greater regime strength, not less.

In rejoinder, it might reasonably be wondered why---if manipulation indicates strength---why protests occur after highly manipulated elections? In my view, this interpretation unnecessarily exports the instrumental logic of protest from the regime-weakness model to the regime-strength model. As discussed earlier in the paper, instrumental calculations about overthrowing incumbents are but one component of a long and varied literature on protest, which has identified multiple other expressive and strategic motives for protest. These other motives may foster protest movements in the aftermath of rigged elections, even when severe electoral manipulation indicates considerable regime strength. Second, if manipulation does indeed indicate strength, incumbents have every incentive to build and deploy those strengths in reaction to stronger opposition movements (who will themselves be more capable of initiating protest). Evidence that incumbents do, indeed, appear to increase manipulation in the face of increased protest threat is presented in the supplementary appendix.

It also helps us make sense of secular patterns that are puzzling if election fraud indicates regime weakness, as others have held. The pattern depicted in Figure \@ref(fig:elex-hist) for example, which shows high levels of manipulation and low levels of protest, is sensible in a world where the more powerful and durable incumbents produce greater levels of manipulation, and where incumbents who struggle to produce high levels of manipulation are ones who face resource scarcity, institutional barriers to impunity for their agents, and defections among erstwhile allies. It also helps explain why some incumbents manufacture their way to highly implausible margins of victory, reasoning that a demonstration of their extra-electoral power may suggest to allies and bystanders that the incumbent may survive protest despite a low natural popularity. Scenarios where incumbents use massive fraud to post margins of victory in the eighty and ninety percentage point range are harder to explain under a worldview where more manipulation indicates deeper regime weakness.

@@

What does all this tell us? First, these findings support and extend the line of research initiated by Simpser [@simpser2013]. Understanding election manipulation as indicative of effective sources of regime strength comports with a growing literature on the principal-agent dynamics of manipulation [@Harvey2019;@rundlett2016deliver;@Harvey2022] and the importance of patronage resources and electoral processes in sustaining authoritarian regimes [@hale_patronal_2014]. This study offers empirical evidence that authoritarian governments who manipulate more are stronger than those who manipulate less (all else equal). It further illustrates how the mechanisms which underpin Simpser's model of manipulation-as-strength---coordination effects among pro-regime agents and costly signaling of incumbent resources---can function even when such signals of strength have failed to 'deter' protest.

Second, these results may inform how researchers think about the causes of election manipulation. They suggest that the degree to which the opposition's capacity to organize street protest can limit election fraud has been overestimated. Consequently, theories of electoral manipulation that emphasize alternatives---such as resource availability [@greene_why_2007], principal-agent problems [@rundlett2016deliver], and legal risks to manipulators [@Harvey2022]---should be further developed, and may yield new policy interventions. In particular, it may change some current understandings of the role of election monitors, courts, and other third parties in upholding election integrity. As noted earlier, prior studies argue that these actors restrain manipulation by revealing information about fraud and intensifying protest risk. However, if more extensive manipulation is associated with more quickly subduing protest movements, it may be that other mechanisms are at work. For instance, it may be that election observation makes it harder for front-line agents to avoid personal consequences for engaging in manipulation [@Harvey2022].

By showing that the infrastructure of fraud can actually reduce the cost of post-election protest to the regime, this result poses a challenge to theory of democratization by election [@Schedler2013;@Morgenbesser2019;@edgellWhenWhereElections2018]. If the capacity to engage in widespread manipulation makes it easier to both tolerate the existence of an opposition and to suppress it during periods of contestation [@dahlPolyarchyParticipationOpposition1971], electoral authoritarianism may be something of an institutional cul-de-sac: the capacity to generate severe election fraud makes it appealing for closed authoritarian regimes to introduce elections, while the manageable risk of protest makes it unappealing to further democratize.

@@

```
-->
Finally, this paper does not examine the role of protest risk in insulating established democracies from democratic backsliding [@fearon_self-enforcing_2011]. This question deserves further investigation, but it is possible to draw a few implications from this study. First, even if protest risk fails to serve as a deterrent in non-democracies, it may still do so in more democratic settings. The likelihood of protest initiation in a democracy may well be higher, triggering at lower levels of fraud and higher incumbent vote-shares, since opposition actors are likely to be better resourced in a liberal democracy (and the government more constrained). At the same time, if the theory described here is accurate, liberal democracies would be constrained against using large-scale election fraud even in the absence of protest risk: they lack the patronage resources and arbitrary enforcement of laws needed for such a scenario.

@@

These results remain consistent under a several alternate specifications presented in an online appendix. The results are supported using two alternative measures of election manipulation, one that covers a broader array of electoral violations, and another that is drawn directly from election observer reports. Additional tests show that incumbents intensify their efforts to manipulate when protest risk is high, that higher capacity for manipulation makes incumbent election loss less likely, and that the small number of successful protests occur across the spectrum of manipulation. This helps rule out an interpretation of the main results in which weaker regimes choose not to manipulate and thus avoid protest.  The results also remain consistent when the selection criteria are relaxed, and a set of electoral democracies are included in the dataset.
